---
title: "A Statistical Investigation of Kyiv Public Transport Distribution"
subtitle: "KSE Probability & Statistics | Summer 2026"
author:
  - Nikita Bezverkhyi
  - Arsen Hanhalo
  - Sofia Tiutiushkina
  - Sofia Danilova
date: "2026-07-13"
lang: en
abstract: |
  This project focuses on researching the Kyiv public transport
  characteristics and planning. The main focus of it is to identify weak
  spots in the existing planning to shed light on the current situation and
  hopefully stimulate improvement. We research the planned intervals
  (headway), the transport capacity and differences between forms of public
  transportation.
format:
  pdf:
    toc: true
    number-sections: true
    # xelatex + шрифти з кирилицею: інакше українські підписи зникають
    # у код-чанках і таблицях (стандартний шрифт pdflatex без кирилиці).
    # macOS: Arial + Menlo. Windows: заміни monofont на "Consolas".
    pdf-engine: xelatex
    mainfont: "Arial"
    monofont: "Menlo"
  html:
    toc: true
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

<!-- ВІДНОВЛЕНО з report.pdf 13.07.2026 + виправлення за код-рев'ю.
     Наратив авторів збережено; виправлено: accepted->rejected (H1/H2/H4),
     числа приведені до рендерених таблиць (H3, H5), codebook, фільтр
     n_departures, підпис Fig. 2, References, Appendix + AI statement.
     ПЕРЕД ЗДАЧЕЮ: кожен автор перечитує свою секцію! -->

**Dataset:** Портал даних Києва, КП «Київпастранс». 1230 route × direction ×
time-slot rows constructed; interval-level analyses (EDA, H1–H2) use the
**1036 rows** with ≥5 departures, categorical/distributional tests (H3, H5)
use the full table. 9 variables.

# Introduction

## Motivation and Context

As in any modern city, in Kyiv public transport is a vital part of a life of
any citizen. And the goal of any city should be to make it as efficient as
possible, because public transit plays a key part of the economic growth of
a city. To put it simply the quicker and more efficient you can get a worker
from point A to point B the more time they will spend working or the more
free time they will have, in turn improving their productivity and mental
state.

Because the practice of publicly available public transport data is quite
rare for Ukrainian cities there are very few researches on this topic. And
those ones which are conducted by city planners are often publicly
unavailable. But we think that the public transportation is a very important
topic for general public and there should be as much available information
on it as possible, and this research is our contribution to that.

## Dataset Description

The data used in this analysis were obtained from the Kyiv Open Data Portal.
The dataset contains plan schedules collected directly from GTFS Static
tables, containing routes, trips, and stop times, spanning a typical active
weekday schedule. Each observation represents a unique route and direction
combined with a specific operational time slot. For the interval-level
analyses (EDA, H1–H2), route–slot combinations with fewer than 5 scheduled
departures were excluded to reduce noise in interval estimates (1230 → 1036
observations); the categorical and distributional tests (H3, H5) use the
full table.

## Codebook

| Variable name | Type | Unit / levels | Description |
|---|---|---|---|
| `route_id` | ID | unique character | Unique identifier for each transport route. |
| `transport_type` | Categorical | Tram / Bus / Trolleybus | Mode of public transport. |
| `time_slot` | Categorical | Peak Morning / Midday / Peak Evening / Night | Categorized operational hours of day. |
| `peak` | Categorical (logical) | TRUE / FALSE | Whether the slot is a rush-hour slot. |
| `n_vehicles` | Quantitative | Count | Number of distinct active vehicles assigned to the route. |
| `total_capacity` | Quantitative | Total passenger count | Combined vehicle passenger capacity. |
| `n_ctrl_points` | Quantitative | Count | Number of schedule control points on a route. |
| `sched_headway_min` | Quantitative | Minutes | Average scheduled interval between runs. |
| `sd_headway_min` | Quantitative | Minutes | Within-slot variability of the interval. |
| `n_departures` | Quantitative | Count | Number of departures in the slot. |

## Research Question and Hypotheses

::: {.callout-important}
**Research question:** Does the city evenly and rationally plan the
allocation of the public transport — by time (day hours), transportation
type and city districts?
:::

The hypotheses we tested:

::: {.callout-note}
$H_1$: Planned intervals during rush hours (7–10, 17–20) are smaller than
non-rush hour intervals.
:::

::: {.callout-note}
$H_2$: Service intervals differ between 4 time slots (evening/night, midday,
morning rush, evening rush) and three transportation types (tram, bus,
trolleybus).
:::

::: {.callout-note}
$H_3$: The transport frequency category (frequent: ≤10 min, average: 10–20
min, rare: >20 min, during rush) does not depend on the transport type
(Tram/Bus/Trolleybus).
:::

::: {.callout-note}
$H_4$: The transport capacity distribution is not balanced around the
district population distribution.
:::

::: {.callout-note}
$H_5$: The service intervals are distributed exponentially (i.e., arrivals
constitute a Poisson process).
:::

# Methodology

## Data Preparation

::: {.callout-important}
1. **Rows dropped:** we dropped all the weekend operations, because they
   would skew our data. Filtered only for the transport that is in use to
   not overestimate the vehicle number. Kept only the stops located in Kyiv,
   because this is the main area of our research. For the interval-level
   analyses (EDA, H1, H2) we additionally excluded route–slot combinations
   with fewer than 5 departures (1230 → 1036 rows), since a mean headway
   based on 2–3 departures is unstable; the categorical and distributional
   tests (H3, H5) use the full 1230-row table to retain all scheduled
   slots — their conclusions are unchanged under the filter.
2. **Missing values:** we handled the missing values by omission to avoid
   the imputation bias. Used `na.rm = TRUE` to skip the missing values
   during aggregation. Because we used `lag()` to calculate the headways,
   the first departures created NA values which were removed.
3. **Variable transformation:** normalized the time, created time intervals,
   standardised the `route_type` codes and `bodyType` strings into readable
   categories.
4. **Outlier handling:** set the timetable limits for values ≤0 and ≥120
   minutes.
:::

The full data-preparation pipeline is executed in
`01_data_preparation.qmd`; the key steps are shown below for transparency
(not re-evaluated here).

### Load the data

```{r}
#| eval: false
library(curl)
urls <- c(
  gtfs_kyiv.zip =
    "https://data.kyivcity.gov.ua/dataset/rozklad-rukhu-miskoho-elektrychnoho-ta-avtomobilnoho-transportu-dep-transport/resource/58f0c3d0-9409-4de9-92c8-de4afa035efd/data/download",
  vehicles_kpt.json =
    "https://data.kyivcity.gov.ua/dataset/vidomosti-pro-transportni-zasoby-iaki-obsluhovuiut-pasazhyrski-avtobusni-troleibusni-t-dep-transport/resource/efad1f37-779d-4d8e-9d1e-3a6b345775d6/data/download",
  vehicles_marshrutky.csv =
    "https://data.kyivcity.gov.ua/dataset/vidomosti-pro-transportni-zasoby-iaki-obsluhovuiut-pasazhyrski-avtobusni-troleibusni-t-dep-transport/resource/a8a2b2c6-d709-45fa-ba5c-a42db698e20f/data/download",
  stops_kpt.json =
    "https://data.kyivcity.gov.ua/dataset/dani-pro-mistse-rozmishchennia-zupynok-miskoho-elektrychnoho-ta-avtomobilnoho-transpor-dep-transport/resource/269ffb2e-dac0-4978-9552-10d29a30724b/data/download"
)
dir.create(here("data", "raw"), recursive = TRUE, showWarnings = FALSE)
for (nm in names(urls))
  curl::curl_download(urls[[nm]], here("data", "raw", nm))
```

### Read the GTFS and weekday timetable

```{r}
#| eval: false
gtfs <- read_gtfs(here("data", "raw", "gtfs_kyiv.zip"))
weekday_services <- gtfs$calendar |> filter(monday == 1) |> pull(service_id)
```

### Headways (planned intervals)

Headways are calculated at the first control point (`stop_sequence == 1` —
verified in file; portal description mistakenly says "from 10"). These are
terminal departure times; differences between consecutive departures =
headways.

::: {.callout-warning}
The time in GTFS can be greater than 24:00 (departures after midnight) — so
we check the NA values after conversion.
:::

```{r}
#| eval: false
dep <- gtfs$stop_times |>
  filter(stop_sequence == 1) |>
  inner_join(gtfs$trips, by = "trip_id") |>
  filter(service_id %in% weekday_services) |>
  mutate(dep_sec = as.numeric(hms(departure_time)))

headways <- dep |>
  arrange(route_id, direction_id, dep_sec) |>
  group_by(route_id, direction_id) |>
  mutate(headway_min = (dep_sec - lag(dep_sec)) / 60) |>
  ungroup() |>
  filter(!is.na(headway_min), headway_min > 0, headway_min < 120) |>
  mutate(
    hour = dep_sec %/% 3600,
    time_slot = case_when(
      hour %in% 7:9   ~ "ранковий пік",
      hour %in% 10:16 ~ "день",
      hour %in% 17:19 ~ "вечірній пік",
      TRUE            ~ "вечір/ніч")) |>
  group_by(route_id, direction_id, time_slot) |>
  summarise(sched_headway_min = mean(headway_min),
            sd_headway_min = sd(headway_min),
            n_departures = n(), .groups = "drop")
```

**Sanity check:** verified 2026-07-12 on 5 routes (all 3 modes): tram 1
(4–9m ✓), tram 11 (16–27m ✓), tram 12 (11–17m — partial: direction 1 gives
~22–24m, likely incomplete feed), trolleybus 16 (16–20m ✓), bus 49 (30m ✓).

### The route attributes

```{r}
#| eval: false
route_types <- gtfs$routes |>
  mutate(transport_type = case_when(
    route_type == 0  ~ "трамвай",
    route_type == 3  ~ "автобус",
    route_type == 11 ~ "тролейбус")) |>
  select(route_id, route_short_name, transport_type)

n_stops <- gtfs$stop_times |>
  group_by(trip_id) |> summarise(n = max(stop_sequence)) |>
  inner_join(gtfs$trips, by = "trip_id") |>
  group_by(route_id) |> summarise(n_ctrl_points = max(n))

veh_by_route <- fromJSON(here("data", "raw", "vehicles_kpt.json")) |>
  as_tibble() |>
  filter(status == "Використовується") |>
  separate_rows(routeId, sep = ",") |>
  mutate(routeId = str_trim(routeId),
         transport_type = case_when(
           str_detect(bodyType, "^Трамвай")   ~ "трамвай",
           str_detect(bodyType, "^Тролейбус") ~ "тролейбус",
           str_detect(bodyType, "^Автобус")   ~ "автобус")) |>
  group_by(routeId, transport_type) |>
  summarise(n_vehicles = n_distinct(uid),
            total_capacity = sum(capacity, na.rm = TRUE), .groups = "drop")

route_types_veh <- route_types |>
  left_join(veh_by_route,
            by = c("route_short_name" = "routeId", "transport_type"))
```

### Final analytical table

```{r}
#| eval: false
final <- headways |>
  left_join(route_types_veh, by = "route_id") |>
  left_join(n_stops, by = "route_id") |>
  mutate(peak = time_slot %in% c("ранковий пік", "вечірній пік"))
write_csv(final, here("data", "processed", "analysis_table.csv"))
```

### Data for H4: transportation capacity by district

Each stop in the city stop registry carries a KATOTTG district code, so no
district polygons are needed; the 10 codes were decoded to district names
(`data/raw/katottg_districts.csv`). GTFS control points are matched to
districts via the nearest registry stop. A route's total capacity is split
between districts proportionally to the share of its control points located
in each district.

```{r}
#| eval: false
stops_kpt <- fromJSON(here("data", "raw", "stops_kpt.json")) |> as_tibble() |>
  filter(str_starts(CATUTTC, "UA80")) |>
  left_join(read_csv(here("data", "raw", "katottg_districts.csv")),
            by = c("CATUTTC" = "katottg"))

g_sf <- gtfs$stops |>
  mutate(across(c(stop_lat, stop_lon), as.numeric)) |>
  st_as_sf(coords = c("stop_lon", "stop_lat"), crs = 4326)
k_sf <- st_as_sf(stops_kpt, coords = c("lon", "lat"), crs = 4326)
gtfs_stops_d <- gtfs$stops |>
  mutate(district = stops_kpt$district[st_nearest_feature(g_sf, k_sf)])

capacity_by_district <- gtfs$stop_times |>
  inner_join(gtfs$trips, by = "trip_id") |>
  distinct(route_id, stop_id) |>
  left_join(gtfs_stops_d |> select(stop_id, district), by = "stop_id") |>
  count(route_id, district) |>
  group_by(route_id) |> mutate(share = n / sum(n)) |> ungroup() |>
  left_join(route_types_veh |> select(route_id, total_capacity),
            by = "route_id") |>
  group_by(district) |>
  summarise(capacity = sum(total_capacity * share, na.rm = TRUE))
write_csv(capacity_by_district,
          here("data", "processed", "capacity_by_district.csv"))
```

## Exploratory Data Analysis

```{r}
#| label: setup
#| echo: false
library(tidyverse); library(here); library(lubridate)
d_full <- read_csv(here("data", "processed", "analysis_table.csv"))
# інтервальні аналізи (EDA, H1-H2): слоти з >= 5 відправленнями
d <- d_full |> filter(n_departures >= 5)
```

```{r}
#| label: eda-summary
d |> group_by(peak) |>
  summarise(n = n(), median = median(sched_headway_min),
            mean = mean(sched_headway_min), sd = sd(sched_headway_min)) |>
  knitr::kable(digits = 2,
               caption = "Summary statistics for the analytical sample.")
```

Well both median and mean waiting time periods are quite big. Also we see
that standard deviation for non-peak time is high. Personally I would not
trust the transport if the median waiting time is 23 minutes at the rush
hour.

```{r}
#| label: eda-hist
#| fig-cap: "The interval distribution is right-skewed with a heavy tail."
ggplot(d, aes(sched_headway_min)) +
  geom_histogram(bins = 40) +
  labs(title = "General distribution of the intervals",
       x = "Interval, min", y = "Num") +
  theme_minimal()
```

```{r}
#| label: eda-box-slot
#| fig-cap: "Intervals by time slot: midday stands out with the largest and most variable intervals."
ggplot(d, aes(time_slot, sched_headway_min)) +
  geom_boxplot() +
  labs(x = "Time slot", y = "Interval, min") +
  theme_minimal()
```

```{r}
#| label: eda-box-type
#| fig-cap: "Intervals by transport type: trams hold the tightest schedules."
ggplot(d, aes(transport_type, sched_headway_min, color = transport_type)) +
  geom_boxplot() +
  labs(x = "Transport type", y = "Interval, min") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Statistical Methods

For this research we used the following statistical methods:

1. **Mann–Whitney U Test, one-sided (H1: Peak vs. Off-Peak).** Compares the
   distribution of a continuous variable (headway) between two independent
   groups (peak/off-peak). The one-sided alternative is used because $H_1$
   is directional (intervals are *smaller* during rush hours).
   *Justification:* Shapiro–Wilk and QQ-plots rejected normality in both
   groups, so the rank-based alternative to the parametric t-test is
   required.
2. **Kruskal–Wallis Test (H2: Temporal Slots & Transport Modes).** Compares
   three or more groups within the Mann–Whitney framework. *Justification:*
   normality is rejected and group variances differ, so a rank-based method
   is used; Holm adjustment was applied in post-hoc pairwise comparisons to
   prevent false positives.
3. **Chi-Square ($\chi^2$) Test of Independence (H3: Mode vs. Frequency).**
   Tests the statistical association between two categorical variables.
   *Justification:* one expected cell frequency was below 5, so we used a
   Monte-Carlo simulated p-value instead of the asymptotic approximation.
4. **Chi-Square ($\chi^2$) Goodness-of-Fit Test (H4: Spatial Equity).**
   Evaluates if a sample distribution matches a theoretical multinomial
   baseline (district population shares). Using it we can identify which
   districts are over-served or under-served by transit.
5. **Kolmogorov–Smirnov Test (H5: Stochastic Modeling).** Compares the
   sample empirical distribution (ECDF) against a theoretical continuous
   CDF. *Justification:* the data is continuous, so no artificial binning is
   needed. Because the rate parameter is estimated directly from the data,
   the p-value acts as an approximation.

# Results

## Main Findings

::: {.callout-note}
$H_1$: **Planned intervals during rush hours (7–10, 17–20) are smaller than
non-rush hour intervals.**
:::

```{r}
#| label: h1-diag
d |> group_by(peak) |>
  summarise(shapiro_p = shapiro.test(sched_headway_min)$p.value) |>
  knitr::kable(caption = "Shapiro-Wilk normality test p-values by peak status.")

par(mfrow = c(1, 2))
qqnorm(d$sched_headway_min[d$peak == TRUE],  main = "Peak")
qqline(d$sched_headway_min[d$peak == TRUE])
qqnorm(d$sched_headway_min[d$peak == FALSE], main = "No-peak")
qqline(d$sched_headway_min[d$peak == FALSE])
par(mfrow = c(1, 1))
```

Both groups deviate clearly from normality (heavy right tails on the
QQ-plots; Shapiro–Wilk p-values far below 0.05), which justifies the
rank-based test instead of the t-test.

```{r}
#| label: h1-test
# One-sided test, because H1 is directional ("smaller during rush").
# Group order in the formula is FALSE (off-peak) then TRUE (peak),
# so alternative = "greater" states: off-peak intervals are stochastically
# GREATER than peak intervals, i.e. peak intervals are smaller.
h1_test <- wilcox.test(sched_headway_min ~ peak, data = d,
                       alternative = "greater")
h1_test
```

**The $H_0$ was rejected in favour of the one-sided alternative**
(p = `r format.pval(h1_test$p.value, digits = 2)`): rush-hour intervals are
systematically smaller. The effect is also visible descriptively — the
median scheduled interval is 22.8 min during rush hours versus 26.6 min
off-peak.

::: {.callout-note}
$H_2$: **Service intervals differ between 4 time slots (evening/night,
midday, morning rush, evening rush) and three transportation types (tram,
bus, trolleybus).**
:::

```{r}
#| label: h2
kruskal.test(sched_headway_min ~ factor(time_slot), data = d)
kruskal.test(sched_headway_min ~ factor(transport_type), data = d)
pairwise.wilcox.test(d$sched_headway_min, d$time_slot,
                     p.adjust.method = "holm")
pairwise.wilcox.test(d$sched_headway_min, d$transport_type,
                     p.adjust.method = "holm")
```

**The $H_0$ was rejected for both factors** — there is a difference in both
time slots and transport types. Here we can see that there are no difference
between evening and evening peak and morning and evening peak. But we have
high difference between day and late evening, day and evening peak and
between day and morning peak. We also have very small difference between
late evening and morning peak, which is the most confusing result we
obtained.

Moreover there is more difference between tram and trolleybus than between
bus and either of other transport types.

```{r}
#| label: h2-slot-summary
d |> group_by(time_slot) |>
  summarise(median = median(sched_headway_min),
            mean = mean(sched_headway_min),
            sd = sd(sched_headway_min)) |>
  knitr::kable(digits = 2, caption = "Interval summary by time slot.")
```

With this summary table we can see the difference between the day and every
other time-slot. Midday has the highest standard deviation — 14.6, median —
29.5 and mean — 32.09 values, while other time-slots have more or less the
same values. So the real gap is not between peak and non-peak but between
the day and everything else. Moreover, higher standard deviation at the
daytime tells us about an unstable schedule at this period of time compared
to the rest of the day.

```{r}
#| label: h2-type-summary
d |> group_by(transport_type) |>
  summarise(median = median(sched_headway_min)) |>
  arrange(median) |>
  knitr::kable(digits = 2, caption = "Median interval by transport type.")
```

Here is a just-to-compare table to see that the tram is the best-planned
transport per our investigation. Nevertheless we should mention there are
way more bus observations than other types of transport (590 bus rows vs.
125 tram and 321 trolleybus rows), and that particular type of transport
needs more attention from the city transport administration.

::: {.callout-note}
$H_3$: **The transport frequency category (frequent: ≤10 min, average: 10–20
min, rare: >20 min, during rush) does not depend on the transport type
(Tram/Bus/Trolleybus).**
:::

```{r}
#| label: h3
# H3 та H5 пораховані на повній таблиці (без фільтра >= 5 відправлень),
# щоб зберегти всі запланові слоти; висновки за фільтром не змінюються
peak_data <- d_full |> filter(peak) |>
  mutate(freq_cat = cut(sched_headway_min, breaks = c(0, 10, 20, Inf),
                        labels = c("частий", "середній", "рідкісний")))
tab3 <- table(peak_data$transport_type, peak_data$freq_cat)
tab3
chi3 <- chisq.test(tab3, simulate.p.value = TRUE, B = 10000)
round(chi3$expected, 2)
chi3
```

**The $H_0$ was rejected**: trams are used more frequently while buses are
used less frequently, but there are a bigger amount of them.

The high chi-square value of `r round(unname(chi3$statistic), 1)` shows us
that the difference in the intervals for the three modes of transportation
across categories is very significant and differs greatly from
proportionally equal values; for if Kyiv had planned the intervals for
buses, trams, and trolleybuses to be strictly the same, the value would have
been close to 0. The expected frequency for trams in the "frequent"
category turned out to be less than the threshold (4.01 < 5). Therefore, the
standard approximation to the chi-square distribution could give us a
misleading value. For this reason, we used a p-value simulation: the
computer independently generates a distribution from 10000 random
combinations of the available data, and the degrees of freedom in this case
are meaningless — therefore, the results show NA. The simulated p-value of
0.000099 is very small and shows us that the chance of such a transportation
distribution occurring by chance is very low.

An analysis of this data confirmed that trams have the highest scheduled
arrival frequency during peak hours; however, since Kyiv's tram network is
smaller than its bus and trolleybus networks, this result does not allow us
to draw broad conclusions about the superiority of one mode of
transportation over another. However, we can draw a specific conclusion that
provides a clear practical recommendation for passengers: if there is a
choice between a bus stop and a tram stop in the same direction, it is
always better to choose the tram.

::: {.callout-note}
$H_4$: **The transport capacity distribution is not balanced around the
district population distribution.**
:::

```{r}
#| label: h4
cap_d <- read_csv(here("data", "processed", "capacity_by_district.csv"))
pop   <- read_csv(here("data", "raw", "population_districts.csv"))

d4 <- inner_join(cap_d, pop, by = "district")
test4 <- chisq.test(x = d4$capacity, p = d4$population / sum(d4$population))
test4
d4 <- d4 |> mutate(stdres = round(test4$stdres, 1)) |> arrange(stdres)
d4 |> knitr::kable(
  caption = "Observed transport capacity, population, and standardized residuals by district.")
```

**The $H_0$ (capacity proportional to population) was rejected**: there is a
two-way disbalance — huge residential districts are lacking public transport
while smaller central districts have an excess of transport volume.

```{r}
#| label: h4-metro
metro <- read_csv(here("data", "processed", "metro_stations_by_district.csv"))
d4m <- d4 |> left_join(metro, by = "district") |>
  mutate(metro_per_100k = metro_stations / population * 1e5)
d4m |> select(district, stdres, metro_stations, metro_per_100k) |>
  knitr::kable(digits = 2,
    caption = "Standardized residuals and metro accessibility by district.")
cor(d4m$stdres, d4m$metro_per_100k, method = "spearman")
```

```{r}
#| label: h4-fig
#| fig-width: 9
#| fig-height: 5.5
#| fig-cap: "Two-way disbalance of planned surface capacity relative to population; metro does not compensate."
d4m |>
  ggplot(aes(x = reorder(district, stdres), y = stdres, fill = stdres > 0)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0("М: ", metro_stations),
                y = ifelse(stdres > 0, stdres + 14, stdres - 14)), size = 3.4) +
  scale_fill_manual(values = c("firebrick", "steelblue")) +
  coord_flip() +
  labs(title = "Дисбаланс планової провізної здатності відносно населення",
       subtitle = "Стандартизовані залишки chi-square тесту: «−» недообслужені, «+» переобслужені.\nМ: n — кількість станцій метро в районі (метро дисбаланс не компенсує)",
       x = NULL, y = "Стандартизований залишок",
       caption = "Дані: Портал даних Києва (GTFS, реєстр ТЗ, схема метро); населення — ГУ статистики у м. Києві, 2022") +
  theme_minimal(base_size = 13)
```

We evaluated whether the spatial distribution of planned surface transport
passenger capacity across Kyiv's districts matches their respective
population distributions (chi-square goodness-of-fit; population serves as a
demand proxy since ticket validation data is not publicly available). The
hypothesis of proportionality was soundly rejected ($\chi^2 = 53{,}820$,
$df = 9$, $p < 2.2 \times 10^{-16}$), revealing a bidirectional imbalance.
Large residential districts are severely underserved — Darnytskyi accounts
for 11.8% of the city's population but receives only 7.1% of surface
capacity ($stdres = -102.4$), followed closely by Sviatoshynskyi ($-65.0$)
and Solomianskyi ($-56.0$). Conversely, the central Shevchenkivskyi
($+155.1$) and Podilskyi ($+113.5$) districts, each housing ~7% of the
population, receive 11–13% of the total capacity.

We tested whether this disparity is offset by the underground rapid transit
network: Spearman's rank correlation between the test residuals and the
number of metro stations per 100,000 residents is a mere 0.08, indicating no
systematic compensation. Sviatoshynskyi and Solomianskyi operate at a
structural deficit across both transit systems; the sole exception is
Pecherskyi, whose surface transit deficit is offset by the highest metro
density in the city. Notably, given a sample size of this magnitude
(hundreds of thousands of passenger slots), even a minor deviation from
proportionality would yield statistical significance; thus, the true
analytical insight lies not in the p-value itself, but in the sign and
magnitude of the standardized residuals.

::: {.callout-note}
$H_5$: **The service intervals are distributed exponentially (i.e., arrivals
constitute a Poisson process).**
:::

```{r}
#| label: h5
# як і H3 — на повній таблиці (див. примітку в Data Preparation)
for (type in c("автобус", "трамвай", "тролейбус")) {
  intervals <- d_full |>
    filter(transport_type == type, peak == TRUE) |>
    pull(sched_headway_min) |> na.omit()
  lambda_est <- 1 / mean(intervals)
  ks_res <- ks.test(intervals, "pexp", rate = lambda_est)
  cat(sprintf("%-10s n = %3d  lambda = %.4f  D = %.4f  p = %.3g\n",
              type, length(intervals), lambda_est,
              ks_res$statistic, ks_res$p.value))
}
```

**The $H_0$ (exponentially distributed intervals, i.e. fully random
arrivals) was rejected for all three modes.** The D-statistic ranges from
0.2337 for trams to 0.3266 for trolleybuses (buses: 0.3157). This value
indicates the discrepancy between the scheduled intervals and the ideal
random curve: if the vehicles arrived at stops in a completely random and
independent manner, the values would be close to 0. The p-value for all
modes is far below 0.05 (the largest is 0.001068 for trams). This shows that
the probability that such a distribution of intervals occurred by chance is
practically zero: the schedules are genuinely planned rather than random.
Buses and trolleybuses show the greatest deviation from randomness,
indicating that fixed intervals are maintained more strictly there. Since we
used data from the transit plan rather than actual arrival times, we cannot
definitively state whether public transit schedules can be trusted in
reality, but we can offer a practical recommendation: planned intervals are
structured and predictable, so checking the schedule is worthwhile.

## Main figure

```{r}
#| label: fig-violin
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Planned headways by time slot and transport mode (weekdays)."
ggplot(d, aes(x = time_slot, y = sched_headway_min, fill = transport_type)) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_boxplot(width = 0.18, outlier.alpha = 0.35, fill = "white") +
  facet_wrap(~transport_type) +
  scale_fill_manual(values = c("автобус" = "#e07a7a",
                               "трамвай" = "#63b56f",
                               "тролейбус" = "#7aa6e0")) +
  guides(fill = "none") +
  labs(title = "Трамвайна мережа тримає найменші інтервали, тоді як автобуси критично просідають удень",
       subtitle = "Аналіз планових інтервалів руху за часовими слотами в будні дні",
       x = "Період доби", y = "Плановий інтервал руху (хвилини)",
       caption = "Джерело: розрахунки на основі GTFS-фідів КП «Київпастранс» (2026)") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Bonus: Plan vs. Reality Pilot

The core of this project analyses the *planned* schedule. As a bonus pilot
we collected 362 real-time GPS snapshots of the whole fleet (Sunday 09:58 —
Monday 09:17, one snapshot every 2 minutes; 164,451 vehicle positions) and
checked, on the same three routes used for methodology validation, how the
actual Monday morning rush compares to the plan. An "arrival" is a vehicle
crossing into a 200 m radius around the route terminal; the inside/outside
state resets only after the vehicle moves 450 m away (hysteresis), so a
vehicle parked at the terminal is not double-counted. The 2-minute polling
grid adds ±2 min of noise per event.

```{r}
#| label: pilot
library(geosphere)   # install.packages("geosphere") if missing
rt <- read_csv(here("data", "processed", "realtime_positions.csv")) |>
  mutate(dt = as_datetime(timestamp, tz = "Europe/Kyiv"))

pilots <- c("1_1", "2_12", "3_100")
trips_g      <- read_csv(here("data", "raw", "gtfs", "trips.txt"))
stop_times_g <- read_csv(here("data", "raw", "gtfs", "stop_times.txt"))
stops_g      <- read_csv(here("data", "raw", "gtfs", "stops.txt"))

# main terminal of each direction = most frequent first control point
terminals <- stop_times_g |>
  filter(stop_sequence == 1) |>
  inner_join(trips_g, by = "trip_id") |>
  filter(route_id %in% pilots) |>
  count(route_id, direction_id, stop_id, sort = TRUE) |>
  group_by(route_id, direction_id) |> slice(1) |> ungroup() |>
  left_join(stops_g, by = "stop_id")

arrivals_at <- function(rt_route, term_lat, term_lon) {
  rt_route |>
    arrange(vehicle_id, timestamp) |>
    mutate(dist = distHaversine(cbind(lon, lat), c(term_lon, term_lat))) |>
    group_by(vehicle_id) |>
    mutate(inside = purrr::accumulate(
             dist, .init = FALSE,
             \(state, dd) if (dd < 200) TRUE else if (dd > 450) FALSE else state)[-1],
           arrival = inside & !lag(inside, default = FALSE)) |>
    ungroup() |> filter(arrival) |> arrange(dt) |> pull(dt)
}

win <- rt |> filter(dt >= ymd_hm("2026-07-13 06:45", tz = "Europe/Kyiv"))

results <- list()
for (i in seq_len(nrow(terminals))) {
  t <- terminals[i, ]
  arr <- arrivals_at(win |> filter(route_id == t$route_id),
                     t$stop_lat, t$stop_lon)
  hw <- diff(as.numeric(arr) / 60)
  hw <- hw[hw > 2 & hw < 90]
  results[[i]] <- tibble(route_id = t$route_id,
                         direction_id = t$direction_id,
                         n_arrivals = length(arr),
                         fact_mean = mean(hw), fact_median = median(hw),
                         headways = list(hw))
}
actual <- bind_rows(results)

plan_pilot <- d |>
  filter(route_id %in% pilots, time_slot == "ранковий пік") |>
  group_by(route_id, direction_id) |>
  summarise(plan_headway = mean(sched_headway_min), .groups = "drop")

actual |>
  left_join(plan_pilot, by = c("route_id", "direction_id")) |>
  select(-headways) |>
  knitr::kable(digits = 1,
    caption = "Actual vs. planned morning-rush headways, Monday 2026-07-13.")
```

```{r}
#| label: pilot-wilcoxon
hw_tram1 <- actual |> filter(route_id == "1_1") |> pull(headways) |> unlist()
plan_tram1 <- plan_pilot |> filter(route_id == "1_1") |>
  summarise(m = mean(plan_headway)) |> pull(m)
wilcox.test(hw_tram1, mu = plan_tram1, alternative = "greater")
```

Trolleybus 16 matched its planned headway almost exactly (median 15.8 min
vs. planned 16.0 at the "vul. Mrii" terminal), while tram 1 ran markedly
less frequently than planned (7.3–7.6 min vs. planned 4.8–5.8; one-sided
Wilcoxon signed-rank against the planned value: p = 0.0035). Bus 49 produced
too few events (2 per direction) for any conclusion. Ties in the Wilcoxon
test stem from the 2-minute time quantization, so the normal approximation
with continuity correction is used.

# Discussion

## Summary

We find a statistically significant, structurally critical difference in
public transport intervals across times of day and modes of transport, and a
pronounced two-way spatial imbalance: large residential districts (above all
Darnytskyi) receive far less planned surface capacity than their population
share, while central districts receive far more — and the metro network does
not compensate for this gap. This directly answers our research question:
Kyiv's planned transport grid is unevenly distributed, showing exceptional
vulnerabilities during off-peak midday operation blocks and in the largest
residential districts.

## Interpretation and Contextualisation

The practical impact of these variations is extensive. The massive shift in
standard deviation up to 14.6 minutes during midday reveals an erratic
supply model that directly harms mid-day commuters. Trams maintain the
tightest, most consistent planning priority across the municipal system
layout. On the spatial side, the residual structure of the H4 test provides
a concrete, data-backed priority list for transit planners: Darnytskyi,
Sviatoshynskyi and Solomianskyi are systematically under-provisioned even
after accounting for metro access.

## Limitations and Self-Critique

1. **Scheduled vs. actual reality:** this project investigates GTFS Static
   parameters, tracking planned intents rather than accounting for actual
   real-time congestion delays.
2. **Demand proxies:** we rely on population distributions as a demand
   baseline, which does not map variable daily commuting footprints; the
   2022 population structure may also differ from the current wartime one.
3. **Coverage:** the GTFS feed covers only the municipal operator
   («Київпастранс»); private minibuses are absent, and known feed gaps exist
   (e.g. missing runs on tram 12, confirmed by citizen requests on the data
   portal).
4. **Capacity accounting:** a vehicle assigned to several routes is counted
   in each of them, so `total_capacity` reflects capacity available to a
   route rather than the unique city fleet.
5. **Temporal scope:** the real-time pilot (see *Bonus: Plan vs. Reality
   Pilot*) suggests the plan is meaningful but not guaranteed: trolleybus 16
   matched its planned morning-rush headway almost exactly, while tram 1 ran
   markedly less frequently than planned (7.3–7.6 min vs. 4.8–5.8 min,
   Wilcoxon p = 0.0035). Still, one morning of data limits
   generalizability, and the 2-minute polling grid adds ±2 min noise per
   event.

## What We Would Do Differently

If we had more time, we would deploy active servers to parse automated daily
scripts for two weeks, capturing true live metrics against static baseline
assumptions.

# References {-}

1. **Розклад руху — GTFS Static.** Портал даних Києва, Департамент
   транспортної інфраструктури / КП «Київпастранс».
   <https://data.kyivcity.gov.ua/dataset/rozklad-rukhu-miskoho-elektrychnoho-ta-avtomobilnoho-transportu-dep-transport>
2. **Відомості про транспортні засоби** (пасажиромісткість, парк ТЗ).
   <https://data.kyivcity.gov.ua/dataset/vidomosti-pro-transportni-zasoby-iaki-obsluhovuiut-pasazhyrski-avtobusni-troleibusni-t-dep-transport>
3. **Зупинки** (координати, коди КАТОТТГ районів).
   <https://data.kyivcity.gov.ua/dataset/dani-pro-mistse-rozmishchennia-zupynok-miskoho-elektrychnoho-ta-avtomobilnoho-transpor-dep-transport>
4. **Схема Київського метрополітену** (ресурс `underground` набору
   «Розклад руху») — використана для розподілу 52 станцій метро по районах.
5. **Чисельність наявного населення по районах м. Києва, 2022.** Головне
   управління статистики у м. Києві.
   <http://www.kyiv.ukrstat.gov.ua/p.php3?c=1123&lang=1>

# Appendix

## Team Contributions

<!-- ПІДТВЕРДИТИ КОМАНДОЮ перед здачею: -->

| Team member | Contribution |
|---|---|
| Arsen Hanhalo | Data pipeline (GTFS, vehicles, stops), headway computation and validation, H4 (district capacity, metro robustness check). |
| Sofia Danilova | Exploratory data analysis, H1 and H2 (diagnostics, Mann–Whitney, Kruskal–Wallis with post-hoc). |
| Sofia Tiutiushkina | H3 (chi-square independence with Monte-Carlo p-value) and H5 (Kolmogorov–Smirnov). |
| Nikita Bezverkhyi | Report assembly and editing, presentation slides, realtime GPS collection support. |

All team members have reviewed the full code and analysis and take
collective responsibility for the report and presentation.

## Use of AI

Code skeletons, project structure and data-source discovery were assisted by
Claude (Anthropic); all code was reviewed, adapted and executed by the team,
and the statistical interpretations and narrative were written by the
authors. AI-assisted code is disclosed here in accordance with the course
policy on generative AI.
